Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-06-29 manuscript/main.tex
unregularized computation streamlined

\begin{definition}[Hybrid inner product between $n$ observations and a single object]
Let $\boldsymbol{\xi} = (\psi, \mathbf{s}) \in \mathcal{H}$ be a hybrid object with $\psi \in \mathcal{F}$ and $\mathbf{s} \in \mathbb{R}^p$, and let $\mathbf{a}^{(1)}, \ldots, \mathbf{a}^{(K)}$ denote the basis coefficients of $\psi$ with respect to basis $\Phi^{(1)}, \ldots, \Phi^{(K)}$. Then the inner product between $\boldsymbol{\xi}$ and the $n$ basis-approximated hybrid observations $\widetilde{\mathbf{W}}_1, \ldots, \widetilde{\mathbf{W}}_n$, concatenated as a column vector, is denoted $\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}} \in \mathbb{R}^n$ and computed as
	\begin{equation*}
		\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}
		:=
		\begin{bmatrix}
			\langle \widetilde{\mathbf{W}}_1, \boldsymbol{\xi} \rangle_{\mathcal{H}} \\
			\vdots \\
			\langle \widetilde{\mathbf{W}}_n, \boldsymbol{\xi} \rangle_{\mathcal{H}}
		\end{bmatrix}
		=
		\mathbf{Z} \mathbf{s}
		+
		\sum_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \boldsymbol{\Phi}^{(k)} \mathbf{a}^{(k)}. 
	\end{equation*}
\end{definition}

@@

For practical computation, 
let us first denote
\begin{equation*}
\mathbf{b} := \mathbf{y}^\top
[ \widetilde{\mathbf{C}} \, \mathbf{J} \, \mathbf{Z} ].
\end{equation*}
Then we obtain $\mathbf{d}$ by solving a linear system
\begin{equation*}
	\mathbf{J}^\ast \mathbf{d} =
	\mathbf{b}^\top.
\end{equation*}  
Then we obtain the coefficient for the PLS direction by normalize it with respect to the basis $B^*[\mathbf{t}]$:
\begin{equation}\label{compute:pls_direction_coeff}
	\begin{bmatrix} \mathbf{a} \\ \mathbf{\theta} \end{bmatrix}
	= \frac{1}{\sqrt{\mathbf{d}^\top \mathbf{J}^\ast \mathbf{d}}} \mathbf{d}.
\end{equation}
The $l$-th  PLS score for the 
$i$th observation is then computed as the inner product  with respect to the basis $B^*[\mathbf{t}]$:
\begin{equation*}
	\widehat{\rho}_{i}
	= 
	[ \widetilde{\mathbf{C}}  \, \mathbf{Z} ] \, \mathbf{J}^\ast \begin{bmatrix} \mathbf{a} \\ \mathbf{\theta} \end{bmatrix}
\end{equation*}
$ \equiv \widehat{\rho}_{il}$ can be obtained as $\widehat{\rho}_i = \langle \mathbf{W}_i, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}$.
we first construct the following matrix:
	\begin{equation}\label{def:V_star}
	\mathbf{V}^\ast 
	:= 
	n^{-2}\begin{bmatrix}
\mathbf{J}
\widetilde{ \mathbf{C} }^\top \mathbf{y}\mathbf{y}^\top \widetilde{ \mathbf{C} } \mathbf{J} 
& 
\mathbf{J} \widetilde{ \mathbf{C} }^\top 
\mathbf{y}
\mathbf{y}^\top 
\mathbf{Z} 
\\
\mathbf{Z}^\top 
\mathbf{y}
\mathbf{y}^\top \widetilde{ \mathbf{C} }
\mathbf{J} 
&   
\mathbf{Z}^\top 
\mathbf{y}
\mathbf{y}^\top 
\mathbf{Z}
	\end{bmatrix} = n^{-2} \mathbf{J}^\ast \widetilde{ \mathbf{C} }^{*\top} \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{ \mathbf{C} }^{*} \mathbf{J}^\ast
	\in \mathbb{R}^{(MK+p) \times (MK+p)}.
\end{equation}

@@

	\subsection{Proof of Propositions \ref{Proposition: generalized rayleigh quotient} and \ref{Proposition: Regularized generalized rayleigh quotient} }\label{section:proof:eigenproblem_derivation}
		Here, $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$ is expressed as $\boldsymbol{\xi}^{*\top} \mathbf{V}^\ast \boldsymbol{\xi}^*$ based on the aforementioned basis approximations. Thus, finding $\boldsymbol{\xi}^*$ that maximizes $\boldsymbol{\xi}^{*\top} \mathbf{V}^\ast \boldsymbol{\xi}^*$ amounts to finding $\widehat{\boldsymbol{\xi}}$ that maximizes the sample squared covariance $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$, which is the first step of the PLS algorithm. The constraint $\boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*=1$ scales the obtained PLS component to have a unit norm, that is, $\langle \widehat{\boldsymbol{\xi}}, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}=1$.

>>>>

\begin{definition}[Hybrid inner product between $n$ observations and a single object]\label{def:n_one_inner_product}
Let $\boldsymbol{\xi} = (\psi, \mathbf{s}) \in \mathcal{H}$ be a hybrid object with $\psi \in \mathcal{F}$ and $\mathbf{s} \in \mathbb{R}^p$, and let $\mathbf{a}^{(1)}, \ldots, \mathbf{a}^{(K)}$ denote the basis coefficients of $\psi$ with respect to basis $\Phi^{(1)}, \ldots, \Phi^{(K)}$. Then the inner product between $\boldsymbol{\xi}$ and the $n$ basis-approximated hybrid observations $\widetilde{\mathbf{W}}_1, \ldots, \widetilde{\mathbf{W}}_n$, concatenated as a column vector, is denoted $\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}} \in \mathbb{R}^n$ and computed as
	\begin{equation*}
		\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}
		:=
		\begin{bmatrix}
			\langle \widetilde{\mathbf{W}}_1, \boldsymbol{\xi} \rangle_{\mathcal{H}} \\
			\vdots \\
			\langle \widetilde{\mathbf{W}}_n, \boldsymbol{\xi} \rangle_{\mathcal{H}}
		\end{bmatrix}
		=
		\mathbf{Z} \mathbf{s}
		+
		\sum_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \boldsymbol{\Phi}^{(k)} \mathbf{a}^{(k)}. 
	\end{equation*}
\end{definition}

@@

For the computation of the estimated PLS direction  $\hat{\boldsymbol{\xi}}$, 
we obtain the basis coefficients of its functional part, denoted by $\mathbf{d}^{(1)}, \ldots, \mathbf{d}^{(K)} \in \mathbb{R}^{M}$,
and the scalar part  
$\mathbf{d} \in \mathbb{R}^p$,
by solving linear systems
\begin{equation*}
 \boldsymbol{\Phi}^{(1)}
\mathbf{d}^{(1)} 
=
(\widetilde{\mathbf{C}}^{(1)} \boldsymbol{\Phi}^{(1)})^\top
\mathbf{y},
\ldots
,
\boldsymbol{\Phi}^{(K)}
\mathbf{d}^{(K)} 
=
(\widetilde{\mathbf{C}}^{(K)} \boldsymbol{\Phi}^{(K)})^\top
\mathbf{y},
~\text{and}~
\mathbf{d} := \mathbf{Z}^\top \mathbf{y} .
\end{equation*}
We then normalize these coefficients and scalar parts by the hybrid Hilbert norm (Definition
\ref{def:hilbert_space}),
which can be computed as
\begin{equation*}
\|	\hat{\boldsymbol{\xi}} \|_{\mathcal{H}} = 
\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}.
\end{equation*}
Therefore, the estimated PLS direction is the tuple $\hat{\boldsymbol{\xi}} \in \mathcal{H}$, defined as
\begin{equation}
	\hat{\boldsymbol{\xi}} :=
	\frac{1}{
	\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}
	}
	\biggl(
	\sum_{m=1}^M d^{(1)}_m \phi^{(1)}_m,
	\ldots,
	\sum_{m=1}^M d^{(K)}_m \phi^{(K)}_m,
	\mathbf{d}
	\biggr).
\end{equation}
Finally, the estimated $l$th PLS scores for the $n$ observations,
denoted $\hat{\boldsymbol{\rho}} \in \mathbb{R}^n$
 is computed simultaneously as the hybrid inner product (Definition \ref{def:n_one_inner_product}):
	\begin{equation*}
	\langle \widetilde{\mathbf{W}}, \hat{\boldsymbol{\xi}} \rangle_{\mathcal{H}}
	=
	\frac{1}{\|	\hat{\boldsymbol{\xi}} \|_{\mathcal{H}}}
	\biggl(
	\mathbf{Z} \mathbf{d}
	+
	\sum_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \boldsymbol{\Phi}^{(k)} \mathbf{d}^{(k)}
	\biggr)
	. 
\end{equation*}

@@

	\subsection{Proof of Propositions \ref{Proposition: generalized rayleigh quotient} and \ref{Proposition: Regularized generalized rayleigh quotient} }\label{section:proof:eigenproblem_derivation}

	we first construct the following matrix:
	\begin{equation}\label{def:V_star}
		\mathbf{V}^\ast 
		:= 
		n^{-2}\begin{bmatrix}
			\mathbf{J}
			\widetilde{ \mathbf{C} }^\top \mathbf{y}\mathbf{y}^\top \widetilde{ \mathbf{C} } \mathbf{J} 
			& 
			\mathbf{J} \widetilde{ \mathbf{C} }^\top 
			\mathbf{y}
			\mathbf{y}^\top 
			\mathbf{Z} 
			\\
			\mathbf{Z}^\top 
			\mathbf{y}
			\mathbf{y}^\top \widetilde{ \mathbf{C} }
			\mathbf{J} 
			&   
			\mathbf{Z}^\top 
			\mathbf{y}
			\mathbf{y}^\top 
			\mathbf{Z}
		\end{bmatrix} = n^{-2} \mathbf{J}^\ast \widetilde{ \mathbf{C} }^{*\top} \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{ \mathbf{C} }^{*} \mathbf{J}^\ast
		\in \mathbb{R}^{(MK+p) \times (MK+p)}.
	\end{equation}

		Here, $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$ is expressed as $\boldsymbol{\xi}^{*\top} \mathbf{V}^\ast \boldsymbol{\xi}^*$ based on the aforementioned basis approximations. Thus, finding $\boldsymbol{\xi}^*$ that maximizes $\boldsymbol{\xi}^{*\top} \mathbf{V}^\ast \boldsymbol{\xi}^*$ amounts to finding $\widehat{\boldsymbol{\xi}}$ that maximizes the sample squared covariance $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$, which is the first step of the PLS algorithm. The constraint $\boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*=1$ scales the obtained PLS component to have a unit norm, that is, $\langle \widehat{\boldsymbol{\xi}}, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}=1$.
